peter_models

### 2024-12-13 peter_models.qmd
updating models, commenting out deprecated models

dataset_clean$Q29_reordered  <- factor(dataset_clean$Q29,
                            levels = c("(4) Not at all interested in politics",
                                     "(3) Not that interested in politics",
                                     "(2) Somewhat interested",
                                     "(1) Very interested in politics"))

dataset_clean$Q271_reordered <- factor(dataset_clean$Q271,
                            levels = c("(5) Not at all important",
                                     "(4) Slightly important",
                                     "(3) Moderately important",
                                     "(2) Very important",
                                     "(1) Extremely important"))

ord_mod <- polr(Q29 ~ Q271_reordered + race_recoded +
                  age_recoded +
            educ +  # Educational Attainment
            S3B + # Gender
            Q21, # Party affiliation
          data = dataset_clean, Hess=TRUE)

summary(ord_mod)

pvals <- pnorm(-abs(summary(ord_mod)$coef[,"t value"]))*2
ctable <- cbind(summary(ord_mod)$coef,pvals)

ctable

(exp_coefs <- exp(cbind(OR=coef(ord_mod),confint(ord_mod))))

head(ord_mod$fitted.values)
head(predict(ord_mod))

confusionMatrix(predict(ord_mod), dataset_clean$Q29, mode="everything")

@@

library(MASS) # package for model
library(caret) # package for confusion matrix

# dataset_clean$Q271_reordered <- factor(dataset_clean$Q271,
#                             levels = c("(5) Not at all important",
#                                      "(4) Slightly important",
#                                      "(3) Moderately important",
#                                      "(2) Very important",
#                                      "(1) Extremely important"))

ord_mod_interactpoliticalID <- polr(Q29_reordered ~ Q271_reordered * Q21 +
                               race_recoded +
                  age_recoded +
            educ +  # Educational Attainment
            S3B, # Gender
          data = dataset_clean, Hess=TRUE)

summary(ord_mod_interactpoliticalID)

pvals_interactpoliticalID <- pnorm(-abs(summary(ord_mod_interactpoliticalID)$coef[,"t value"]))*2
ctable_interactpoliticalID <- cbind(summary(ord_mod_interactpoliticalID)$coef,pvals_interactpoliticalID)

ctable_interactpoliticalID

(exp_coefs_interact <- exp(cbind(OR=coef(ord_mod_interactpoliticalID),confint(ord_mod_interactpoliticalID))))

head(ord_mod_interactpoliticalID$fitted.values)
head(predict(ord_mod_interactpoliticalID))

confusionMatrix(predict(ord_mod_interactpoliticalID), dataset_clean$Q29, mode="everything")

@@

# Load packages
library(nnet)  # for multinomial logistic regression
library(tidyverse)
library(car)
library(effects)

# ensure outcome variable is a factor but not ordered
dataset_clean$Q29_unordered <- factor(dataset_clean$Q29, ordered = FALSE)
dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)

# Fit multinomial model
m_mult <- multinom(Q29_unordered ~ Q271 * S3B + 
                     race_recoded + 
                     age_recoded +
            educ + # Educational Attainment
            Q21, # Party affiliation
          data = dataset_clean)

# Calculate predicted probabilities of marginal effects
# Create sequence of Q271_simple values
x_vals <- seq(min(dataset_clean$Q271_simple), max(dataset_clean$Q271_simple), length.out = 100)

# Function to calculate average predicted probabilities
get_avg_probs <- function(model, newdata, var_name) {
    preds <- predict(model, newdata = newdata, type = "probs")
    colMeans(preds)
}

# Calculate marginal effects by race
races <- unique(dataset_clean$race_recoded)
marg_effects <- list()
# 
for(race in races) {
    probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
    for(i in seq_along(x_vals)) {
        temp_data <- dataset_clean
        temp_data$Q271_simple <- x_vals[i]
        temp_data$race_major <- race
        probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
    }
    marg_effects[[race]] <- data.frame(
        x = x_vals,
        race = race,
        prob1 = probs[,1],
        prob2 = probs[,2],
        prob3 = probs[,3],
        prob4 = probs[,4]
    )
}

# Combine all marginal effects
marg_effects_df <- bind_rows(marg_effects)

# Create plots for marginal effects
# Reshape data for plotting
marg_effects_long <- marg_effects_df %>%
    pivot_longer(cols = starts_with("prob"),
                names_to = "outcome",
                values_to = "probability") %>%
    mutate(outcome = factor(outcome,
                          levels = paste0("prob", 1:4),
                          labels = levels(dataset_clean$Q29_unordered)))

# Plot marginal effects
p1 <- ggplot(marg_effects_long,
             aes(x = x, y = probability, color = outcome)) +
    geom_line() +
    facet_wrap(~race) +
    theme_minimal() +
    labs(title = "Marginal Effects of Racial Identity Importance by Race",
         x = "Importance of Racial Identity",
         y = "Predicted Probability",
         color = "Political Interest Level") +
    theme(legend.position = "bottom")
print(p1)

# Model fit statistics and comparisons
print("Model summaries:")
print(summary(m_mult))

# Calculate McFadden's R-squared
null_model <- multinom(Q29_unordered ~ 1, data = dataset_clean)
mcfadden_r2 <- 1 - logLik(m_mult)/logLik(null_model)

print("McFadden's R-squared:")
print(mcfadden_r2)

# Classification accuracy
pred_class <- predict(m_mult)
conf_matrix <- table(pred_class, dataset_clean$Q29_unordered)
accuracy <- sum(diag(conf_matrix))/sum(conf_matrix)

print("Classification accuracy:")
print(accuracy)
print("Confusion Matrix:")
print(conf_matrix)
# 
# Print AIC
print(AIC(m_mult))

# Calculate predicted probabilities
pred_data <- expand.grid(
    Q271 = unique(dataset_clean$Q271),
    age_recoded = unique(dataset_clean$age_recoded),
    race_recoded = unique(dataset_clean$race_recoded),
    educ = unique(dataset_clean$educ),
    S3B = unique(dataset_clean$S3B),
    Q21 = unique(dataset_clean$Q21)
)
# 
# Get predictions
preds <- predict(m_mult, newdata = pred_data, type = "probs")
pred_data <- cbind(pred_data, as.data.frame(preds))

# Create separate plots for each outcome level
pred_long <- pred_data %>%
    pivot_longer(cols = starts_with("("),
                names_to = "interest_level",
                values_to = "probability")

# Plot for each outcome level
p2 <- ggplot(pred_long,
             aes(x = Q271, y = probability, color = race_recoded)) +
    geom_line() +
    facet_wrap(~interest_level) +
    theme_minimal() +
    labs(title = "Predicted Probabilities by Outcome Level",
         x = "Importance of Racial Identity",
         y = "Probability",
         color = "Race/Ethnicity") +
    theme(legend.position = "bottom")
print(p2)

@@

```{r}
# generate predictions for the ordinal model and the ordinal model  with interactions
predict(ord_mod, new_dataset, type="probs")
predict(ord_mod_interactions, new_dataset, type="probs")

new_dataset <- data.frame(race_recoded=c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA"),
                          Q271=c("(5) Not at all important", "(4) Slightly important", "(3) Moderately important", "(2) Very important", "(1) Extremely important", "(4) Slightly important", "(5) Not at all important"),
                          age_recoded = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+", "18-29"),
                          S3B = c("(1) Man", "(2) Woman","(3) Non-binary","(4) Something else (Specify)", "(1) Man", "(2) Woman", "(1) Man"),
                          educ = c("No college experience", "Some college experience", "College graduate", "No college experience", "Some college experience", "College graduate", "College graduate"),
                          Q21 = c("(1) Republican",  "(2) Democrat","(3) Independent", "(4) Other party", "(1) Republican",  "(2) Democrat",  "(3) Independent"))

predict(m_mult, new_dataset, type = "probs")
```

@@

# Confusion matrix for multinomial model
conf_matrix
```

>>>>

```{r}
# library(MASS) # package for model
# library(caret) # package for confusion matrix
# 
# dataset_clean$Q29_reordered  <- factor(dataset_clean$Q29,
#                             levels = c("(4) Not at all interested in politics",
#                                      "(3) Not that interested in politics",
#                                      "(2) Somewhat interested",
#                                      "(1) Very interested in politics"))
# 
# 
# dataset_clean$Q271_reordered <- factor(dataset_clean$Q271,
#                             levels = c("(5) Not at all important",
#                                      "(4) Slightly important",
#                                      "(3) Moderately important",
#                                      "(2) Very important",
#                                      "(1) Extremely important"))
# 
# ord_mod <- polr(Q29 ~ Q271_reordered + race_recoded +
#                   age_recoded +
#             educ +  # Educational Attainment
#             S3B + # Gender
#             Q21, # Party affiliation
#           data = dataset_clean, Hess=TRUE)
# 
# summary(ord_mod)
# 
# 
# pvals <- pnorm(-abs(summary(ord_mod)$coef[,"t value"]))*2
# ctable <- cbind(summary(ord_mod)$coef,pvals)
# 
# ctable
# 
# (exp_coefs <- exp(cbind(OR=coef(ord_mod),confint(ord_mod))))
# 
# head(ord_mod$fitted.values)
# head(predict(ord_mod))
# 
# confusionMatrix(predict(ord_mod), dataset_clean$Q29, mode="everything")

@@

# library(MASS) # package for model
# library(caret) # package for confusion matrix
# 
# # dataset_clean$Q271_reordered <- factor(dataset_clean$Q271,
# #                             levels = c("(5) Not at all important",
# #                                      "(4) Slightly important",
# #                                      "(3) Moderately important",
# #                                      "(2) Very important",
# #                                      "(1) Extremely important"))
# 
# ord_mod_interactpoliticalID <- polr(Q29_reordered ~ Q271_reordered * Q21 +
#                                race_recoded +
#                   age_recoded +
#             educ +  # Educational Attainment
#             S3B, # Gender
#           data = dataset_clean, Hess=TRUE)
# 
# 
# 
# summary(ord_mod_interactpoliticalID)
# 
# 
# pvals_interactpoliticalID <- pnorm(-abs(summary(ord_mod_interactpoliticalID)$coef[,"t value"]))*2
# ctable_interactpoliticalID <- cbind(summary(ord_mod_interactpoliticalID)$coef,pvals_interactpoliticalID)
# 
# ctable_interactpoliticalID
# 
# (exp_coefs_interact <- exp(cbind(OR=coef(ord_mod_interactpoliticalID),confint(ord_mod_interactpoliticalID))))
# 
# head(ord_mod_interactpoliticalID$fitted.values)
# head(predict(ord_mod_interactpoliticalID))
# 
# confusionMatrix(predict(ord_mod_interactpoliticalID), dataset_clean$Q29, mode="everything")

@@

# 
# # Load packages
# library(nnet)  # for multinomial logistic regression
# library(tidyverse)
# library(car)
# library(effects)
# 
# 
# # ensure outcome variable is a factor but not ordered
# dataset_clean$Q29_unordered <- factor(dataset_clean$Q29, ordered = FALSE)
# dataset_clean$Q271_simple <- as.numeric(dataset_clean$Q271)
# 
# # Fit multinomial model
# m_mult <- multinom(Q29_unordered ~ Q271 * S3B + 
#                      race_recoded + 
#                      age_recoded +
#             educ + # Educational Attainment
#             Q21, # Party affiliation
#           data = dataset_clean)
# 
# 
# 
# # Calculate predicted probabilities of marginal effects
# # Create sequence of Q271_simple values
# x_vals <- seq(min(dataset_clean$Q271_simple), max(dataset_clean$Q271_simple), length.out = 100)
# 
# # Function to calculate average predicted probabilities
# get_avg_probs <- function(model, newdata, var_name) {
#     preds <- predict(model, newdata = newdata, type = "probs")
#     colMeans(preds)
# }
# 
# # Calculate marginal effects by race
# races <- unique(dataset_clean$race_recoded)
# marg_effects <- list()
# # 
# for(race in races) {
#     probs <- matrix(0, nrow = length(x_vals), ncol = length(levels(dataset_clean$Q29_unordered)))
#     for(i in seq_along(x_vals)) {
#         temp_data <- dataset_clean
#         temp_data$Q271_simple <- x_vals[i]
#         temp_data$race_major <- race
#         probs[i,] <- get_avg_probs(m_mult, temp_data, "Q271_simple")
#     }
#     marg_effects[[race]] <- data.frame(
#         x = x_vals,
#         race = race,
#         prob1 = probs[,1],
#         prob2 = probs[,2],
#         prob3 = probs[,3],
#         prob4 = probs[,4]
#     )
# }
# 
# # Combine all marginal effects
# marg_effects_df <- bind_rows(marg_effects)
# 
# # Create plots for marginal effects
# # Reshape data for plotting
# marg_effects_long <- marg_effects_df %>%
#     pivot_longer(cols = starts_with("prob"),
#                 names_to = "outcome",
#                 values_to = "probability") %>%
#     mutate(outcome = factor(outcome,
#                           levels = paste0("prob", 1:4),
#                           labels = levels(dataset_clean$Q29_unordered)))
# 
# # Plot marginal effects
# p1 <- ggplot(marg_effects_long,
#              aes(x = x, y = probability, color = outcome)) +
#     geom_line() +
#     facet_wrap(~race) +
#     theme_minimal() +
#     labs(title = "Marginal Effects of Racial Identity Importance by Race",
#          x = "Importance of Racial Identity",
#          y = "Predicted Probability",
#          color = "Political Interest Level") +
#     theme(legend.position = "bottom")
# print(p1)
# 
# # Model fit statistics and comparisons
# print("Model summaries:")
# print(summary(m_mult))
# 
# # Calculate McFadden's R-squared
# null_model <- multinom(Q29_unordered ~ 1, data = dataset_clean)
# mcfadden_r2 <- 1 - logLik(m_mult)/logLik(null_model)
# 
# print("McFadden's R-squared:")
# print(mcfadden_r2)
# 
# # Classification accuracy
# pred_class <- predict(m_mult)
# conf_matrix <- table(pred_class, dataset_clean$Q29_unordered)
# accuracy <- sum(diag(conf_matrix))/sum(conf_matrix)
# 
# print("Classification accuracy:")
# print(accuracy)
# print("Confusion Matrix:")
# print(conf_matrix)
# # 
# # Print AIC
# print(AIC(m_mult))
# 
# # Calculate predicted probabilities
# pred_data <- expand.grid(
#     Q271 = unique(dataset_clean$Q271),
#     age_recoded = unique(dataset_clean$age_recoded),
#     race_recoded = unique(dataset_clean$race_recoded),
#     educ = unique(dataset_clean$educ),
#     S3B = unique(dataset_clean$S3B),
#     Q21 = unique(dataset_clean$Q21)
# )
# # 
# # Get predictions
# preds <- predict(m_mult, newdata = pred_data, type = "probs")
# pred_data <- cbind(pred_data, as.data.frame(preds))
# 
# # Create separate plots for each outcome level
# pred_long <- pred_data %>%
#     pivot_longer(cols = starts_with("("),
#                 names_to = "interest_level",
#                 values_to = "probability")
# 
# 
# # Plot for each outcome level
# p2 <- ggplot(pred_long,
#              aes(x = Q271, y = probability, color = race_recoded)) +
#     geom_line() +
#     facet_wrap(~interest_level) +
#     theme_minimal() +
#     labs(title = "Predicted Probabilities by Outcome Level",
#          x = "Importance of Racial Identity",
#          y = "Probability",
#          color = "Race/Ethnicity") +
#     theme(legend.position = "bottom")
# print(p2)

@@

```{r}
# generate predictions for the ordinal model and the ordinal model  with interactions
# predict(ord_mod, new_dataset, type="probs")
predict(ord_mod_interactions, new_dataset, type="probs")

# new_dataset <- data.frame(race_recoded=c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA"),
#                           Q271=c("(5) Not at all important", "(4) Slightly important", "(3) Moderately important", "(2) Very important", "(1) Extremely important", "(4) Slightly important", "(5) Not at all important"),
#                           age_recoded = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+", "18-29"),
#                           S3B = c("(1) Man", "(2) Woman","(3) Non-binary","(4) Something else (Specify)", "(1) Man", "(2) Woman", "(1) Man"),
#                           educ = c("No college experience", "Some college experience", "College graduate", "No college experience", "Some college experience", "College graduate", "College graduate"),
#                           Q21 = c("(1) Republican",  "(2) Democrat","(3) Independent", "(4) Other party", "(1) Republican",  "(2) Democrat",  "(3) Independent"))

# predict(m_mult, new_dataset, type = "probs")
```

@@

# Confusion matrix for multinomial model
